### 2022-12-29 paper.Rmd
editing

When optimizing a bicriterion objective such as $SSE_{kplus}$, the ideal outcome would be to find a partition where both the means and the variances are very similar between groups, i.e., a which partition where both the standard $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ are close to the global optimum value. Generally, research on multi-objective optimization has shown that cases are rare where a single solution satisfies all criteria to an optimal degree [@ferligoj1992direct]. Usually, a tradeoff is necessary. However, a motivating example illustrates that aiming to simultaneously optimize similarity of means and variances is not only desirable, but may actually be feasible. I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatterplot of the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ across partitions. Note that for both criteria, a larger value is better, i.e., indicates higher similarity with regard to means and variances, respectively. Therefore, to obtain overall between-group similarity, a partition should be chosen that is depicted in the upper right corner of the scatterplot. However, optimal k-means anticlustering would create two groups that are not particularly similar with regard to their variance. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $\mathit{SSE_{Var}}$. Similarly, the partition that maximizes $\mathit{SSE_{Var}}$ (the triangular symbol in Figure 1) is not well-suited to optimize $\mathit{SSE}$ at the same time. The diamond symbol represents the partition that maximizes (the unweighted) $SSE_{kplus}$, i.e., the sum of $\mathit{SSE}$ and $\mathit{SSE_{Var}}$. This partition is located in the right upper corner and yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $\mathit{SSE_{Var}}$.

```{r, fig.cap = "The relationship between the objectives $SSE$ and $SSE_{Var}$ across all ways to partition $N = 14$ normally distributed random values into $K = 2$ groups. The partition represented by the triangular minimizes the difference in variances between groups; the partition represented by the square minimizes the difference in means between groups; the partition represented by the diamond maximizes $SSE_{kplus}$ and represents a good tradeoff, satisfying both criteria to an almost optimal degree.", fig.height = 5, fig.width = 5}

# Plot SSE vs SSE(VAR)
plot(
  df[, c("kmeans", "kvar")], 
  pch = 4, col = "darkgrey", 
  las = 1, cex = .5,
  xlab = "SSE",
  ylab = "SSE_Var"
)

@@

The outlined logic to equalize the variance, which is the second moment of a distribution, can be generalized to higher order moments in a straight forward manner. The *j*'th sample moment of a variable $x = (x_1, \ldots, x_N)$ can be computed as

@@

The present paper focused on the new k-plus criterion for anticlustering, but only superficially considered the multicriterion nature of this criterion: A simple unweighted sum approach---combined with a generic local maximum neighbourhood search---was used to optimize the k-plus objective. More sophisticated multicriterion approaches may yield improved results, as well as the opportunity to deliberately select among partitionings that may be preferred according to different k-plus objectives. @brusco2019 showed how a direct algorithm can be used to optimize multiple criteria at the same time in anticlustering applications. Their study focused on approximating the pareto set of efficient solutions that optimize diversity and dispersion, which is essential in applications that require high within-group heterogeneity. Future research should investigate if k-plus anticlustering may profit from similar advanced algorithmic treatment.

This paper's focus on the k-plus objective only instead of optimization algorithms followed two reasonings. First, the applications showed that the local maximum search---maximizing the unweighted k-plus sum---yields satisfying results in practice that probably suffice in most applications. Second, it was of interest to investigate the merits of the k-plus objective function per se, independently from the algorithm that is used to optimize it. To obtain a fair assessment in the simulation study, it was therefore important that each anticlustering objective was optimized using the same algorithm. Differences in performance cannot be interpreted if the objective function as well as the algorithm used to optimize it differs between methods. It was of particular interest to compare the k-plus objective with the popular diversity objective. @papenberg2020 compared k-means anticlustering and diversity anticlustering[^clusterediting] in a simulation study and concluded that the diversity criterion should be preferred because k-means does not equalize the spread of the distribution between groups. In contrast, maximizing the diversity leads to an appropriate balance between location and spread. Using the k-plus extension, however, the k-means criterion has regained attractivity: When opting for similarity in means and variances---and for increasing $N$, skewness and kurtosis as well---, k-plus anticlustering clearly outperformed diversity anticlustering.

>>>>

When optimizing a bicriterion objective such as $\mathit{SSE_{kplus}}$, the ideal outcome would be to find a partition where both the means and the variances are very similar between groups, i.e., a which partition where both the standard $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ are close to the global optimum value. However, research on multi-objective optimization has shown that cases are rare where a single solution satisfies all criteria to an optimal degree, making a tradeoff necessary [@ferligoj1992direct]. A motivating example illustrates that in the case of anticlustering, aiming to simultaneously optimize similarity of means and variances may however be feasible. I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatterplot of the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ across partitions. To divide the 14 elements into two groups that are similar with regard to their mean and variance, a partition should be chosen that is depicted in the upper right corner of the scatterplot. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $\mathit{SSE_{Var}}$. Similarly, the partition that maximizes $\mathit{SSE_{Var}}$ (the triangular symbol in Figure 1) is not well-suited to optimize $\mathit{SSE}$ at the same time. The diamond symbol represents the partition that maximizes $\mathit{SSE_{kplus}}$, i.e., the sum of $\mathit{SSE}$ and $\mathit{SSE_{Var}}$. This partition is located in the right upper corner and yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $\mathit{SSE_{Var}}$. Thus, the optimal partition according to $\mathit{SSE_{kplus}}$ fulfills the two constituting criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ to an astonishing degree.

```{r, fig.cap = "The relationship between the objectives $SSE$ and $SSE_{Var}$ across all ways to partition $N = 14$ normally distributed random values into $K = 2$ groups. The partition represented by the triangular minimizes the difference in variances between groups; the partition represented by the square minimizes the difference in means between groups; the partition represented by the diamond maximizes $SSE_{kplus}$, and satisfying both criteria to an almost optimal degree.", fig.height = 5, fig.width = 5}

# Plot SSE vs SSE(VAR)
plot(
  df[, c("kmeans", "kvar")], 
  pch = 4, col = "darkgrey", 
  las = 1, cex = .5,
  xlab = "",
  ylab = ""
)
mtext(side = 2, text = expression(italic(SSE[Var])), line = 3)
mtext(side = 1, text = expression(italic(SSE)), line = 3)

@@

The k-plus approach to equalizing the variance between groups can be generalized to higher order moments in a straight forward manner. The *j*'th sample moment of a variable $x = (x_1, \ldots, x_N)$ can be computed as

@@

The present paper focused on the new k-plus criterion for anticlustering, but only superficially considered the multicriterion nature of this criterion: A simple unweighted sum approach---combined with a generic local maximum neighbourhood search---was used to optimize the k-plus objective. The sole focus on the objective function followed two reasonings. First, the applications showed that the local maximum search, maximizing the unweighted k-plus sum, yielded satisfying results that probably suffice in most applications. It seems that k-plus anticlustering has a fortunate solution space where many suitable partitions exist that fulfill multiple criteria at the same time (see Figure 1). Still, more sophisticated approaches may yield improved results, as well as the opportunity to deliberately select among partitionings that may be preferred according to different k-plus objectives. @brusco2019 gave an example of how direct algorithms can be employed to optimize multiple criteria at the same time in anticlustering applications aiming for high within-group heterogeneity. Future research should investigate if k-plus anticlustering may profit from similar algorithmic treatment. Second, it was of interest to investigate the merits of the k-plus objective function per se, independently from the algorithm that is used to optimize it. To obtain a fair assessment in the simulation study, it was therefore important that each anticlustering objective was optimized using the same algorithm. Differences in performance cannot be interpreted if the objective function as well as the algorithm used to optimize it differs between methods. It was of particular interest to compare the k-plus objective with the popular diversity objective. @papenberg2020 compared k-means anticlustering and diversity anticlustering[^clusterediting] in a simulation study and concluded that the diversity criterion should be preferred because k-means does not equalize the spread of the distribution between groups. In contrast, maximizing the diversity leads to an appropriate balance between location and spread. Using the k-plus extension, however, the k-means criterion has regained attractivity: When opting for similarity in means and variances---and for increasing $N$, skewness and kurtosis as well---, k-plus anticlustering clearly outperformed diversity anticlustering.
